Push-Based Signal-Function Functional Reactive Programming (Thesis Proposal)

### 2012-05-21 Docs/Proposal/abstract.tex
final proposal version

\section*{Abstract}
Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first class objects in the
program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
leads to inefficient and wasteful evaluation of the FRP system.

>>>>

\section*{Abstract}
Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
functional languages, but has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
direct access to the time-varying  values by the programmer, but rather provides {\em signal functions} as first-class objects in the
program. All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation, producing output from
the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function FRP systems, but
leads to inefficient and wasteful evaluation of the FRP system.

### 2012-05-21 Docs/Proposal/background.tex
final proposal version

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
as output and/or take functions as input. We say that functions are {\em first class values} in a language.

Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
or object based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.

@@

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.

@@

The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
spawned to evaluate the new behavior. Unfortunately, The implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
used, and leads to thread leakage when one of the merged events is itself the merging of other events.

@@

In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for Events to form a functor. To construct our example application,
we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
a function which takes a behavior for audio data, lowpass frequency, and highpass frequency, and produces a behavior for the filtered audio data. This
function would be written in terms of the integration primitive common to most FRP libraries. Figure~\ref{figure:classic-frp-example} provides an example.

@@

An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
which are themselves not first class. This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
implementing FRP with first-class signals and events.

@@

In the example for classic FRP (Figure~\ref{figure:classic-frp-example}), we note pure values which are not referentially transparent,
since they depend on real-world values. While notation in this manner is appealing, the behavior and event streams must be implemented either
by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
IO actions to be treated as pure values, as in most other classic FRP systems. Further, note that event streams, in particular, must be 
explicitly split in order to avoid time leaks and violation of causality. Finally, our program is not modular, since it is not parameterized on
its inputs, and for the same reason, it cannot be suspended as a state value without reference to these input sources.

>>>>

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produces functions
as output and/or take functions as input. We say that functions are {\em first-class values} in a language.

Most reactive programs are written in an imperative style, using low-level and non-composable abstractions including callbacks
or object-based event handlers, called by event loops. This ties the model of interactivity to low-level implementation details such as timing and event handling models.

@@

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in functional reactive
models should of course be efficiently executable. This has proven to be the main challenge in implementing Functional Reactive Programming.

@@

The Reactive~\cite{Elliott2009} system is a push-based FRP system with first-class behaviors and events. The primary insight of Reactive is
the separation of reactivity (or changes in response to events whose occurrence time could not be know beforehand) from time-dependence. This
gives rise to {\em reactive normal form}, which represents a behavior as a constant or simple time-varying value, with an event stream carrying values
which are also behaviors in reactive normal form. Push-based evaluation is achieved by forking a Haskell thread to evaluate the head behavior,
while waiting on the evaluation of the event stream. Upon an event occurrence, the current behavior thread is killed and a new thread
spawned to evaluate the new behavior. Unfortunately, the implmentation of Reactive uses a tenuous technique which depends on also forking threads to evaluate
two haskell values concurrently in order to implement event merging. This relies on the library author to ensure consistency when this technique is
used, and leads to thread leakage when one of the merged events is itself the merging of other events.

@@

In classic FRP systems, it is common for the Behavior type to form an applicative functor, and for the Event type to form a functor. To construct our example application,
we lift the exponential function, with a tuning constant, to a behavior, which we then apply to the $x$ and $y$ values of the mouse. Finally, we assume
a function which takes a behavior for audio data, lowpass frequency, and highpass frequency, and produces a behavior for the filtered audio data. This
function would be written in terms of the integration primitive common to most FRP libraries. Figure~\ref{figure:classic-frp-example} provides an example.

@@

An alternative approach to FRP was first proposed in work on Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs. This library
utilized the abstraction of Arrows~\cite{Hughes2000} to structure {\em signal functions}. Arrows are abstract type constructors with input and output type
parameters, together with sequential ({\tt >>>}) and parallel ({\tt first} and {\tt second}) composition, branching ({\tt ***}), and lifting ({\tt arr}) functions
to the Arrow type. Signal functions are the first-class abstraction in this approach, they represent time-dependent and reactive transformers of events and signals, 
which are themselves not first class, since such values cannot be directly manipulated by the programmer.
This approach has two motivations: it increases modularity since both the input and output of signal functions may be transformed,
as opposed to signals or events which may only be transformed in their output, and it avoids a large class of time and space leaks which have emerged when
implementing FRP with first-class signals and events.

@@

In the example for classic FRP (Figure~\ref{figure:classic-frp-example}), we note pure values which are not referentially transparent,
since they depend on real-world values. While notation in this manner is appealing, the behavior and event streams must be implemented either
by ``compiling'' to a dependency graph with continuations for sampling, as in reactive-banana, or by using backdoors which allow
IO actions to be treated as pure values, as in most other classic FRP systems.
Further, note that event streams, in particular, must be explicitly split in order to avoid time leaks and violation of causality.
Finally, our program is not modular, since it is not parameterized on its inputs, and for the same reason,it cannot be suspended as a state value without reference
to these input sources. This requires evaluation to take place in a single event loop.

### 2012-05-21 Docs/Proposal/introduction.tex
final proposal version

The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
{\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
representing them implicitly, signal functions prohibit accessing past or future time values, only values in the closure of the signal function
and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
signal function and acts on the output values. This function is elided in existing literature on signal functions. 
A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.

@@

To address both challenges in a single FRP implementation, it seems desirable to combine the approaches, creating a push-based
signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors\footnote{Behaviors are generally
referred to as {\em signals} in signal function literature. This is unfortunate, since a signal function may operate on events, signals, or some combination of the
two.}, provides a representation which suggests a method of push-based evaluation. In previous implementations of FRP, events were simply
option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
the truly continuous values which must be repeatedly sampled.

N-ary FRP provides a model of a type system for a push-based signal-function FRP system. However, in order to represent signal functions
as continuations on collections of inputs rather than on single values, we represent the signal vectors as {\em phantom types}~\cite{Cheney03} over which
the signal function type is parameterized for both inputs and outputs. From these phantom types, we can type indices into the signal vector, which
contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,
which forms the basis for evaluating only those parts of the network which depend on the updated values.

>>>>

The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
{\em transformers} of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
signal function and acts on the output values. This evaluation function is elided in existing literature on signal functions. 
A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.

@@

To address both challenges in a FRP implementation, it seems desirable to combine the approaches, creating a push-based
signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
recent work on N-ary FRP, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
a method of push-based evaluation. In previous implementations of FRP, events were simply
option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
the truly continuous values which must be repeatedly sampled.

N-ary FRP provides a model of a type system for a push-based signal-function FRP system. In order to represent signal functions
as continuations on collections of inputs rather than on single values, we represent the signal vectors as types over which
the signal function type is parameterized for both inputs and outputs. Indices into the signal vector are typed with the signal vector. These indices
contain the actual input or output values and form elements for a collection of inputs or outputs. Obviously the collection need not have all possible elements,
which forms the basis for evaluating only those parts of the network which depend on the updated values.

### 2012-05-21 Docs/Proposal/proposed_work.tex
final proposal version

Recent work on types for signal-function FRP systems has yielded N-ary FRP~\cite{Sculthorpe2011}, which types signal function inputs and outputs
with {\em signal vectors} rather than types of values. Signal vectors are phantom types: signal or event type constructor taking a value type,
or a product type of signal vectors. This permits individual components of a signal function's input or output to be designated as an event
or as a signal, and provides precisely the flexibility necessary to create a push-based implementation.

>>>>

Recent work on types for signal-function FRP systems has yielded N-ary FRP~\cite{Sculthorpe2011}, which types signal function inputs and outputs
with {\em signal vectors} rather than types of values. Signal vectors are types: a signal or event type constructor taking a value type,
or a product type of signal vectors. A signal function then uses signal vectors for its input and output types. Since the actual values which are supplied to the
representation of a signal function also parameterize over signal vectors, this representation permits us to restrict the type of values which may be supplied to the signal
function, even though no values take the type of the signal vector as their whole type. This permits individual components of a signal function's input or output to be
designated as an event or as a signal, and provides precisely the flexibility necessary to create a push-based implementation.
